Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Clustering and Parameter Estimation

### 2020-11-18 summary.Rmd
minor rewording

*Proof (sketch)*. 
Let $P$ and $A$ be organized by community such that the elements of blocks 
$P^{(kl)}$ and $A^{(kl)}$ correspond to the edges between communities $k$ and 
$l$.

@@

In these simulations, Algorithm 2 is compared against sparse subspace 
clustering \cite{soltanolkotabi2014} \cite{noroozi2019estimation} and 
modularity maximization \cite{307cbeb9b1be48299388437423d94bf1} \cite{igraph}, 
and Algorithm 4 is compared against an MLE-based estimator 
\cite{noroozi2019estimation} \cite{307cbeb9b1be48299388437423d94bf1}.

Both Algorithm 2 and sparse subspace clustering produce an affinity matrix $B$ 
for which $B_{ij} \to 0$ if vertices $i$ and $j$ belong in different 
communities. Spectral clustering was performed on $B$ to obtain the final 
community estimates. More specifically, clustering was performed on a 
normalized Laplacian eigenmap of dimension $K + 1$ by fitting Gaussian Mixture 
Model.

For estimating $\{\lambda^{(kl)}\}_K$, both methods use the ground truth 
community labels.

### Sparse subspace clustering and modularity maximization

**TODO** find citations for these

For each simulation, the popularity vectors $\{\lambda^{(kl)}\}_K$ are drawn 
from a joint distribution, the edge probability matrix $P$ is constructed using 
the popularity vectors, and finally an unweighted and undirected adjacency 
matrix $A$ is drawn from $P$. Three community detection methods are applied to 
$A$: Algorithm 2 from this paper, sparse subspace clustering, and modularity 
maximization.

Sparse subspace clustering is described by the following:

\begin{algorithm}[H]
  \DontPrintSemicolon
  \SetAlgoLined
  \caption{Sparse subspace clustering.}
  \KwData{Adjacency matrix $A$, number of communities $K$.}
  \KwResult{Community assignments $1, ..., K$.}
  Embed $A$ in $d$ dimensions using ASE to obtain 
  $\hat{Y} \in \mathbb{R}^{n \times d}$.\;
  Fit a sparse linear regression model to each row of $\hat{Y}$ using the 
  remaining $n - 1$ rows of $Y$ as predictors. The models must exclude the 
  intercept term.\;
  Construct $n \times n$ symmetric matrix $C$ such that $c_{ij}$ is the $j^{th}$
  coefficient for the $i^{th}$ model.\;
  Normalize the columns of $|C|$ to obtain $\tilde{C}$.\;
  Compute the affinity matrix $B = \tilde{C} + \tilde{C}^\top$.\;
  Construct graph $G$ using $B$ as its similarity matrix.\;
  Partition $G$ into $K$ disconnected components (e.g., using edge thresholding 
  or spectral clustering).\;
  Map each partition to the community labels $1, ..., K$\;
\end{algorithm}

For modularity maximization, \citet{307cbeb9b1be48299388437423d94bf1} used the 
Extreme Points (EP) algorithm \cite{le2016}, which is specific to the case 
$K = 2$. In our simulation studies, we used the EP algorithm for $K = 2$ and 
the Louvain algorithm for $K > 2$. The Louvain algorithm is described by the 
following:

\begin{algorithm}[H]
  \DontPrintSemicolon
  \SetAlgoLined
  \caption{Modularity maximization (Louvain Method)}
  \KwData{Adjacency matrix $A$, number of communities $K$.}
  \KwResult{Community assignments $1, ..., K$}
  Initialize the clustering by assigning to each vertex a unique community 
  label.\;
  For each vertex $v_i$, reassign the label to match one of its neighboring 
  vertices, and compute the change in modularity for each reassignment.\;
  Reassign $v_i$ to the community that results in the most positive change in 
  modularity, if possible. If modularity does not increase by changing the 
  label, skip the vertex.\;
  Group vertices with the same label and treat as one vertex.\;
  Repeat steps (3) and (4) until termination or the desired number of 
  communities is estimated.
\end{algorithm}

>>>>

*Proof*. 
Let $P$ and $A$ be organized by community such that the elements of blocks 
$P^{(kl)}$ and $A^{(kl)}$ correspond to the edges between communities $k$ and 
$l$.

@@

For each simulation, the popularity vectors $\{\lambda^{(kl)}\}_K$ are drawn 
from a joint distribution, the edge probability matrix $P$ is constructed using 
the popularity vectors, and finally an unweighted and undirected adjacency 
matrix $A$ is drawn from $P$. Algorithm 2 is then used for community detection, 
and this method is compared against sparse subspace clustering 
\cite{soltanolkotabi2014} \cite{noroozi2019estimation} and modularity 
maximization \cite{307cbeb9b1be48299388437423d94bf1} \cite{igraph}. True 
community labels are used with Algorithm 4 to estimate the popularity 
vectors, and this method is then compared against an MLE-based estimator 
described in \citet{noroozi2019estimation} and
\citet{307cbeb9b1be48299388437423d94bf1}.

Sparse subspace clustering is described as follows. First, an embedding of the 
affinity matrix $A$ is required, and for this, we use spectral decomposition 
using the first $K (K + 1) / 2$ and last $K (K - 1) / 2$ eigenvectors. Then a 
sparse linear regression model is fit to each row of the embedding data matrix 
using the remaining rows as predictors, the rows of which are stored in 
matrix $C$. Finally, an affinity matrix $B = |C| + |C^\top|$ is constructed, 
where $|\cdot|$ is applied entry-wise. If the data in the embedding lie 
perfectly in $K$ subspaces, then $B_{ij} = 0$ for $i, j$ not in the same 
subspace. If the embedding of $P$ is used, then this property holds, but it 
is true in general for the embedding of $A$, in which case a final clustering 
step is required. For these simulations, spectral clustering with the 
normalized Laplacian and Gaussian mixture models is used for this final step.

Modularity maximization is NP-hard, so \citet{307cbeb9b1be48299388437423d94bf1} 
used the Extreme Points (EP) algorithm \cite{le2016}, which is $O(n^{K - 1})$. 
For these simulations, the EP algorithm was used for $K = 2$, and for $K > 2$, 
the Louvain algorithm \cite{Blondel_2008} was used instead.
